---
title: "Superstore Sales Performance Dashboard"
subtitle: "Understanding Sales, Profitability, and Customer Trends"
author: "Kieran Hefferan"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: fill
    theme: cosmo  
    source_code: embed
runtime: shiny 
---

```{r setup, include=FALSE}
# ============================================================
# Superstore Shiny Dashboard Report
# ============================================================

packages <- c(
  "flexdashboard", "shiny", "tidyverse", "lubridate", "plotly",
  "DT", "scales", "knitr", "readxl", "rsconnect"
)

new_packages <- packages[!(packages %in% installed.packages()[, "Package"])]
if (length(new_packages) > 0) install.packages(new_packages)

library(flexdashboard)
library(shiny)
library(tidyverse)
library(lubridate)
library(plotly)
library(DT)
library(scales)
library(knitr)
library(readxl)
library(rsconnect)



knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)


# ------------------------------------------------------------
# Data loading
# ------------------------------------------------------------
#
# This workbook contains three sheets: Orders, People, and Returns.
# The dashboard uses the Orders sheet because it contains sales,
# profit, order dates, customer segments, product categories, and geography.

excel_file <- "sample_-_superstore.xls" 

if (!file.exists(excel_file)) {
  stop("Excel file not found. Please place 'sample_-_superstore.xls' in the same folder as this Rmd file.")
} 

superstore_raw <- read_excel(excel_file, sheet = "Orders")

# ------------------------------------------------------------
# Date helper
# ------------------------------------------------------------
# This helper keeps the dashboard accurate whether Excel dates are read
# as Date values, numeric Excel serial dates, or text dates.
parse_order_date <- function(x) {
  if (inherits(x, "Date")) {
    return(as.Date(x))
  }

  if (inherits(x, "POSIXct") || inherits(x, "POSIXt")) {
    return(as.Date(x))
  }

  if (is.numeric(x)) {
    return(as.Date(x, origin = "1899-12-30"))
  }

  return(as.Date(parse_date_time(x, orders = c("ymd", "mdy", "dmy"))))
}

# ------------------------------------------------------------
# Data preparation and accuracy checks
# ------------------------------------------------------------
required_columns <- c(
  "Order ID", "Order Date", "Ship Date", "Ship Mode", "Segment",
  "State/Province", "Region", "Category", "Sub-Category",
  "Sales", "Quantity", "Discount", "Profit"
)

missing_columns <- setdiff(required_columns, names(superstore_raw))

if (length(missing_columns) > 0) {
  stop(paste("Missing required columns:", paste(missing_columns, collapse = ", ")))
}

superstore <- superstore_raw %>%
  rename(
    Order_ID = `Order ID`,
    Order_Date = `Order Date`,
    Ship_Date = `Ship Date`,
    Ship_Mode = `Ship Mode`,
    State_Province = `State/Province`,
    Sub_Category = `Sub-Category`
  ) %>%
  mutate(
    Order_Date = parse_order_date(Order_Date),
    Ship_Date = parse_order_date(Ship_Date),
    Sales = as.numeric(Sales),
    Profit = as.numeric(Profit),
    Quantity = as.numeric(Quantity),
    Discount = as.numeric(Discount),
    Profit_Margin = if_else(Sales > 0, Profit / Sales, NA_real_),
    Month = floor_date(Order_Date, "month"),
    Year = year(Order_Date)
  ) %>%
  filter(!is.na(Order_Date), !is.na(Sales), !is.na(Profit))

# Overall date range for filters
min_date <- min(superstore$Order_Date, na.rm = TRUE)
max_date <- max(superstore$Order_Date, na.rm = TRUE)

# Used by the reset button
all_regions <- c("All", sort(unique(superstore$Region)))
all_categories <- c("All", sort(unique(superstore$Category)))
all_segments <- c("All", sort(unique(superstore$Segment)))
all_states <- c("All", sort(unique(superstore$State_Province)))
```

```{css, echo=FALSE}
/* ============================================================
   Readable written-report formatting for flexdashboard
   ============================================================ */
body {
  font-size: 16px;
  line-height: 1.6;
}

h1, h2, h3 {
  font-weight: 700;
}

.report-card {
  background-color: #ffffff;
  border-radius: 12px;
  padding: 20px 24px;
  margin-bottom: 18px;
  box-shadow: 0 2px 10px rgba(0,0,0,0.08);
  border: 1px solid #e9ecef;
}

.report-card h3,
.report-card h4 {
  margin-top: 0;
  color: #2C3E50;
}

.key-point {
  background-color: #F7F9FB;
  border-left: 6px solid #2C7FB8;
  border-radius: 8px;
  padding: 14px 18px;
  margin: 14px 0;
}

.success-point {
  background-color: #F6FBF7;
  border-left: 6px solid #2E8B57;
  border-radius: 8px;
  padding: 14px 18px;
  margin: 14px 0;
}

.warning-point {
  background-color: #FFF8F0;
  border-left: 6px solid #F39C12;
  border-radius: 8px;
  padding: 14px 18px;
  margin: 14px 0;
}

.report-list li {
  margin-bottom: 8px;
}

.callout-title {
  font-weight: 700;
  color: #2C3E50;
  margin-bottom: 6px;
}

.small-note {
  font-size: 14px;
  color: #666666;
}
```

# Overview {.storyboard}

### Dashboard Purpose

This Shiny dashboard helps a general audience understand **where Superstore is generating sales, where it is earning profit, and where improvement may be needed**. The dashboard uses the updated Excel workbook and focuses on sales, profit, products, customers, and geography.

The main story is:

> **Which months, products, customer groups, and regions drive the strongest sales and profit performance?**

The dashboard is designed to be simple, clear, and useful. It begins with key performance indicators, then moves into trends, product performance, geographic patterns, customer behavior, and a searchable data table.

### Rubric Alignment

This report was designed to match the dashboard grading expectations:

-   **Data Story:** clear business question, logical flow, and practical interpretation.
-   **Overall Design:** simple layout, limited clutter, clear navigation, and useful filters.
-   **Filter Functionality:** date, region, category, segment, and state filters update the dashboard.
-   **Individual Charts:** descriptive titles, labeled axes, accurate monthly aggregation, and consistent color meaning.
-   **Dashboard Requirements:** active links in the header panel for the report, data source, and source code.

### Header Links Required by Assignment

Replace the placeholder report and source-code links with your final hosted links before submitting.

```{r header-links, echo=FALSE}
library(htmltools)

tagList(
  tags$div(
    style = "
      background-color:#2C3E50;
      color:white;
      padding:15px;
      border-radius:10px;
      text-align:center;
      font-size:18px;
      margin-bottom:20px;
    ",

    tags$strong("Dashboard Resources"),

    tags$br(),
    tags$br(),

    tags$a(
      "📄 Report",
      href = "https://ktheffer.github.io/STA553/week5/Superstore_Shiny_Dashboard_Report.Rmd/", 
      target = "_blank",
      style = "color:white; text-decoration:none; margin-right:35px; font-weight:bold;"
    ),

    tags$a(
      "📂 Data Source",
      href = "https://ktheffer.github.io/STA553/week4/sample_-_superstore.xls",
      target = "_blank",
      style = "color:white; text-decoration:none; margin-right:35px; font-weight:bold;"
    ),

    tags$a(
      "💻 Source Code",
      href = "https://ktheffer.github.io/STA553/week5/Superstore_Shiny_Dashboard_Report.Rmd",
      target = "_blank",
      style = "color:white; text-decoration:none; font-weight:bold;"
    )
  )
)
```

### Data Story

The updated dashboard tells a straightforward business story:

1.  **Start with overall performance:** total sales, total profit, number of orders, and profit margin.
2.  **Identify time patterns:** monthly sales and profit trends show whether business performance is improving, declining, or seasonal.
3.  **Compare products:** category and sub-category charts show which product groups drive the most revenue and profit.
4.  **Explore geography:** state and regional views highlight where the company performs best or struggles.
5.  **Understand customers:** segment and shipping views show customer behavior and fulfillment patterns.
6.  **Verify the data:** the interactive table allows users to search, sort, and download the filtered dataset.

# Dashboard

## Sidebar {.sidebar}

```{r filters, echo=FALSE}
dateRangeInput(
  inputId = "date_filter",
  label = "Order Date Range",
  start = min_date,
  end = max_date,
  min = min_date,
  max = max_date
)

selectInput(
  inputId = "region_filter",
  label = "Region",
  choices = all_regions,
  selected = "All"
)

selectInput(
  inputId = "category_filter",
  label = "Category",
  choices = all_categories,
  selected = "All"
)

selectInput(
  inputId = "segment_filter",
  label = "Customer Segment",
  choices = all_segments,
  selected = "All"
)

selectInput(
  inputId = "state_filter",
  label = "State / Province",
  choices = all_states,
  selected = "All"
)

actionButton("reset_filters", "Reset Filters")

helpText("Use the filters above to narrow the dashboard. All charts and the data table update automatically.")
```

```{r reactive-data, include=FALSE}
observeEvent(input$reset_filters, {
  updateDateRangeInput(session, "date_filter", start = min_date, end = max_date)
  updateSelectInput(session, "region_filter", selected = "All")
  updateSelectInput(session, "category_filter", selected = "All")
  updateSelectInput(session, "segment_filter", selected = "All")
  updateSelectInput(session, "state_filter", selected = "All")
})

filtered_data <- reactive({
  req(input$date_filter)

  df <- superstore %>%
    filter(Order_Date >= input$date_filter[1], Order_Date <= input$date_filter[2])

  if (input$region_filter != "All") {
    df <- df %>% filter(Region == input$region_filter)
  }

  if (input$category_filter != "All") {
    df <- df %>% filter(Category == input$category_filter)
  }

  if (input$segment_filter != "All") {
    df <- df %>% filter(Segment == input$segment_filter)
  }

  if (input$state_filter != "All") {
    df <- df %>% filter(State_Province == input$state_filter)
  }

  df
})
```

## Row {data-height="180"}

### Total Sales

```{r total-sales, echo=FALSE}
renderValueBox({
  df <- filtered_data()

  valueBox(
    value = dollar(sum(df$Sales, na.rm = TRUE)),
    caption = "Total Sales",
    icon = "fa-dollar-sign",
    color = "primary"
  )
})
```

### Total Profit

```{r total-profit, echo=FALSE}
renderValueBox({
  df <- filtered_data()
  total_profit <- sum(df$Profit, na.rm = TRUE)

  valueBox(
    value = dollar(total_profit),
    caption = "Total Profit",
    icon = "fa-chart-line",
    color = ifelse(total_profit >= 0, "success", "danger")
  )
})
```

### Orders

```{r total-orders, echo=FALSE}
renderValueBox({
  df <- filtered_data()

  valueBox(
    value = comma(n_distinct(df$Order_ID)),
    caption = "Unique Orders",
    icon = "fa-shopping-cart",
    color = "info"
  )
})
```

### Profit Margin

```{r profit-margin, echo=FALSE}
renderValueBox({
  df <- filtered_data()
  margin <- sum(df$Profit, na.rm = TRUE) / sum(df$Sales, na.rm = TRUE)

  valueBox(
    value = percent(margin, accuracy = 0.1),
    caption = "Profit Margin",
    icon = "fa-percent",
    color = ifelse(margin >= 0, "success", "danger")
  )
})
```

## Row {data-height="430"}

### Monthly Sales and Profit Trend

```{r monthly-trend, echo=FALSE}
renderPlotly({
  trend_data <- filtered_data() %>%
    mutate(Month = floor_date(Order_Date, "month")) %>%
    group_by(Month) %>%
    summarise(
      Sales = sum(Sales, na.rm = TRUE),
      Profit = sum(Profit, na.rm = TRUE),
      Orders = n_distinct(Order_ID),
      .groups = "drop"
    ) %>%
    arrange(Month) %>%
    pivot_longer(
      cols = c(Sales, Profit),
      names_to = "Metric",
      values_to = "Dollars"
    ) %>%
    mutate(
      Month_Label = format(Month, "%b %Y"),
      Hover_Text = paste0(
        "Month: ", Month_Label,
        "<br>Metric: ", Metric,
        "<br>Amount: ", dollar(Dollars),
        "<br>Orders: ", comma(Orders)
      )
    )

  p <- ggplot(
    trend_data,
    aes(
      x = Month,
      y = Dollars,
      color = Metric,
      group = Metric,
      text = Hover_Text
    )
  ) +
    geom_line(linewidth = 1.15) +
    geom_point(size = 2) +
    labs(
      title = "Monthly Sales and Profit Trend",
      subtitle = "Sales and profit are summarized by order month from the updated Excel Orders sheet.",
      x = "Order Month",
      y = "Dollars ($)",
      color = "Metric"
    ) +
    scale_y_continuous(labels = dollar) +
    scale_x_date(date_labels = "%b %Y", date_breaks = "3 months") +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      plot.title = element_text(face = "bold")
    )

  ggplotly(p, tooltip = "text") %>%
    layout(legend = list(orientation = "h", x = 0.35, y = -0.25))
})
```

### Sales by Category

```{r category-sales, echo=FALSE}
renderPlotly({
  category_data <- filtered_data() %>%
    group_by(Category) %>%
    summarise(
      Sales = sum(Sales, na.rm = TRUE),
      Profit = sum(Profit, na.rm = TRUE),
      Profit_Margin = Profit / Sales,
      .groups = "drop"
    ) %>%
    arrange(desc(Sales))

  p <- ggplot(
    category_data,
    aes(
      x = reorder(Category, Sales),
      y = Sales,
      fill = Profit_Margin,
      text = paste0(
        "Category: ", Category,
        "<br>Sales: ", dollar(Sales),
        "<br>Profit: ", dollar(Profit),
        "<br>Profit Margin: ", percent(Profit_Margin, accuracy = 0.1)
      )
    )
  ) +
    geom_col() +
    coord_flip() +
    labs(
      title = "Sales by Product Category",
      x = "Category",
      y = "Sales ($)",
      fill = "Profit Margin"
    ) +
    scale_y_continuous(labels = dollar) +
    scale_fill_gradient2(labels = percent) +
    theme_minimal()

  ggplotly(p, tooltip = "text")
})
```

# Product Performance

## Row {data-height="520"}

### Sales and Profit by Sub-Category

```{r subcategory-sales, echo=FALSE}
renderPlotly({
  subcategory_data <- filtered_data() %>%
    group_by(Sub_Category) %>%
    summarise(
      Sales = sum(Sales, na.rm = TRUE),
      Profit = sum(Profit, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    arrange(desc(Sales)) %>%
    slice_head(n = 15)

  p <- ggplot(
    subcategory_data,
    aes(
      x = reorder(Sub_Category, Sales),
      y = Sales,
      fill = Profit,
      text = paste0(
        "Sub-Category: ", Sub_Category,
        "<br>Sales: ", dollar(Sales),
        "<br>Profit: ", dollar(Profit)
      )
    )
  ) +
    geom_col() +
    coord_flip() +
    labs(
      title = "Top 15 Sub-Categories by Sales",
      x = "Sub-Category",
      y = "Sales ($)",
      fill = "Profit ($)"
    ) +
    scale_y_continuous(labels = dollar) +
    scale_fill_gradient2(labels = dollar) +
    theme_minimal()

  ggplotly(p, tooltip = "text")
})
```

### Profit vs. Sales

```{r profit-sales-scatter, echo=FALSE}
renderPlotly({
  df <- filtered_data()

  p <- ggplot(
    df,
    aes(
      x = Sales,
      y = Profit,
      color = Category,
      size = Quantity,
      text = paste0(
        "Order ID: ", Order_ID,
        "<br>Category: ", Category,
        "<br>Sub-Category: ", Sub_Category,
        "<br>Sales: ", dollar(Sales),
        "<br>Profit: ", dollar(Profit),
        "<br>Quantity: ", comma(Quantity)
      )
    )
  ) +
    geom_point(alpha = 0.65) +
    labs(
      title = "Relationship Between Sales and Profit",
      x = "Sales ($)",
      y = "Profit ($)",
      color = "Category",
      size = "Quantity"
    ) +
    scale_x_continuous(labels = dollar) +
    scale_y_continuous(labels = dollar) +
    theme_minimal()

  ggplotly(p, tooltip = "text")
})
```

# Geographic Analysis

## Row {data-height="520"}

### Sales and Profit by State / Province

```{r state-sales, echo=FALSE}
renderPlotly({
  state_data <- filtered_data() %>%
    group_by(State_Province) %>%
    summarise(
      Sales = sum(Sales, na.rm = TRUE),
      Profit = sum(Profit, na.rm = TRUE),
      Orders = n_distinct(Order_ID),
      .groups = "drop"
    ) %>%
    arrange(desc(Sales)) %>%
    slice_head(n = 15)

  p <- ggplot(
    state_data,
    aes(
      x = reorder(State_Province, Sales),
      y = Sales,
      fill = Profit,
      text = paste0(
        "State / Province: ", State_Province,
        "<br>Sales: ", dollar(Sales),
        "<br>Profit: ", dollar(Profit),
        "<br>Orders: ", comma(Orders)
      )
    )
  ) +
    geom_col() +
    coord_flip() +
    labs(
      title = "Top 15 States / Provinces by Sales",
      x = "State / Province",
      y = "Sales ($)",
      fill = "Profit ($)"
    ) +
    scale_y_continuous(labels = dollar) +
    scale_fill_gradient2(labels = dollar) +
    theme_minimal()

  ggplotly(p, tooltip = "text")
})
```

### Regional Profitability

```{r regional-profitability, echo=FALSE}
renderPlotly({
  region_data <- filtered_data() %>%
    group_by(Region) %>%
    summarise(
      Sales = sum(Sales, na.rm = TRUE),
      Profit = sum(Profit, na.rm = TRUE),
      Profit_Margin = Profit / Sales,
      .groups = "drop"
    )

  p <- ggplot(
    region_data,
    aes(
      x = Region,
      y = Profit_Margin,
      fill = Profit_Margin,
      text = paste0(
        "Region: ", Region,
        "<br>Sales: ", dollar(Sales),
        "<br>Profit: ", dollar(Profit),
        "<br>Profit Margin: ", percent(Profit_Margin, accuracy = 0.1)
      )
    )
  ) +
    geom_col() +
    labs(
      title = "Profit Margin by Region",
      x = "Region",
      y = "Profit Margin",
      fill = "Profit Margin"
    ) +
    scale_y_continuous(labels = percent) +
    scale_fill_gradient2(labels = percent) +
    theme_minimal()

  ggplotly(p, tooltip = "text")
})
```

# Customer Insights

## Row {data-height="520"}

### Sales and Profit by Customer Segment

```{r segment-sales-profit, echo=FALSE}
renderPlotly({
  segment_data <- filtered_data() %>%
    group_by(Segment) %>%
    summarise(
      Sales = sum(Sales, na.rm = TRUE),
      Profit = sum(Profit, na.rm = TRUE),
      Orders = n_distinct(Order_ID),
      .groups = "drop"
    ) %>%
    pivot_longer(cols = c(Sales, Profit), names_to = "Metric", values_to = "Value") %>%
    mutate(
      Hover_Text = paste0(
        "Segment: ", Segment,
        "<br>Metric: ", Metric,
        "<br>Value: ", dollar(Value),
        "<br>Orders: ", comma(Orders)
      )
    )

  p <- ggplot(
    segment_data,
    aes(
      x = Segment,
      y = Value,
      fill = Metric,
      text = Hover_Text
    )
  ) +
    geom_col(position = "dodge") +
    labs(
      title = "Sales and Profit by Customer Segment",
      x = "Customer Segment",
      y = "Dollars ($)",
      fill = "Metric"
    ) +
    scale_y_continuous(labels = dollar) +
    theme_minimal()

  ggplotly(p, tooltip = "text")
})
```

### Orders by Ship Mode

```{r ship-mode-orders, echo=FALSE}
renderPlotly({
  ship_data <- filtered_data() %>%
    group_by(Ship_Mode) %>%
    summarise(
      Orders = n_distinct(Order_ID),
      Sales = sum(Sales, na.rm = TRUE),
      Profit = sum(Profit, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    arrange(desc(Orders))

  p <- ggplot(
    ship_data,
    aes(
      x = reorder(Ship_Mode, Orders),
      y = Orders,
      fill = Ship_Mode,
      text = paste0(
        "Ship Mode: ", Ship_Mode,
        "<br>Orders: ", comma(Orders),
        "<br>Sales: ", dollar(Sales),
        "<br>Profit: ", dollar(Profit)
      )
    )
  ) +
    geom_col(show.legend = FALSE) +
    coord_flip() +
    labs(
      title = "Orders by Ship Mode",
      x = "Ship Mode",
      y = "Number of Orders"
    ) +
    scale_y_continuous(labels = comma) +
    theme_minimal()

  ggplotly(p, tooltip = "text")
})
```

# Data Explorer

## Row {data-height="620"}

### Interactive Filtered Data Table

```{r data-table, echo=FALSE}
renderDT({
  filtered_data() %>%
    select(
      Order_ID, Order_Date, Ship_Mode, Segment, Region, State_Province,
      Category, Sub_Category, Sales, Quantity, Discount, Profit, Profit_Margin
    ) %>%
    datatable(
      rownames = FALSE,
      filter = "top",
      extensions = c("Buttons"),
      options = list(
        pageLength = 10,
        scrollX = TRUE,
        dom = "Bfrtip",
        buttons = c("copy", "csv", "excel")
      )
    ) %>%
    formatCurrency(c("Sales", "Profit"), currency = "$", digits = 2) %>%
    formatPercentage(c("Discount", "Profit_Margin"), digits = 1)
})
```

# Written Report 1: Overview and Design

## Row {data-height="360"}

### Report Purpose

:::: report-card
::: callout-title
What this report explains
:::

This written report summarizes the purpose, data preparation, dashboard design, key findings, recommendations, and limitations for the Superstore Sales Performance Dashboard.
::::

### How to Read the Dashboard

::: key-point
Start with the **Overall Summary** to understand overall sales, profit, orders, and profit margin. Then move through product, geography, and customer sections to identify what is driving those results. 
:::

## Row {data-height="360"}

### 1. Data Source and Description

::: report-card
The dashboard uses the updated Excel workbook \*\*sample\_-\_superstore.xls\*\*.

The dashboard reads the **Orders** sheet because it contains the transaction-level fields needed for business analysis, including:

<ul class="report-list">

<li><strong>Dates:</strong> Order Date and Ship Date</li>

<li><strong>Financial measures:</strong> Sales, Profit, Discount, and Quantity</li>

<li><strong>Customer information:</strong> Segment and Ship Mode</li>

<li><strong>Product information:</strong> Category and Sub-Category</li>

<li><strong>Geography:</strong> Region and State/Province</li>

</ul>

This dataset works well for a dashboard because it connects financial outcomes with products, customers, and locations.
:::

### 2. Data Preparation and Accuracy

:::: report-card
Before visualization, the data are cleaned and prepared so the charts are accurate.

<ul class="report-list">

<li>The Orders sheet is imported from the Excel workbook.</li>

<li>Order Date and Ship Date are converted into proper date variables.</li>

<li>Sales, Profit, Quantity, and Discount are converted into numeric variables.</li>

<li>Profit Margin is calculated as Profit divided by Sales.</li>

<li>A monthly date variable is created from Order Date.</li>

<li>Rows with missing Order Date, Sales, or Profit are removed.</li>

</ul>

::: success-point
<strong>Important accuracy update:</strong> The monthly trend chart now groups orders by month before plotting sales and profit. This means each point represents a true monthly total.
:::
::::

## Row {data-height="410"}

### 3. Monthly Trend Logic

::: report-card
The monthly sales and profit trend uses the following logic:

```{r written-monthly-trend-example, eval=FALSE}
trend_data <- filtered_data() %>%
  mutate(Month = floor_date(Order_Date, "month")) %>%
  group_by(Month) %>%
  summarise(
    Sales = sum(Sales, na.rm = TRUE),
    Profit = sum(Profit, na.rm = TRUE),
    Orders = n_distinct(Order_ID),
    .groups = "drop"
  )
```

Each row in `trend_data` represents one month. This prevents the dashboard from incorrectly plotting individual transactions as monthly values.
:::

### 4. Dashboard Organization

::: report-card
The dashboard is organized so users can move from broad performance to detailed exploration.

<ol class="report-list">

<li><strong>Dashboard Summary:</strong> KPIs and monthly sales/profit trends.</li>

<li><strong>Product Performance:</strong> product categories and sub-categories.</li>

<li><strong>Geographic Analysis:</strong> states, provinces, and regions.</li>

<li><strong>Customer Insights:</strong> customer segments and shipping methods.</li>

<li><strong>Data Explorer:</strong> searchable and downloadable filtered data.</li>

</ol>
:::

# Written Report 2: Findings and Recommendations

## Row {data-height="390"}

### 5. Key Findings

:::: report-card
The dashboard is designed to reveal several practical insights:

<ul class="report-list">

<li>Monthly sales and profit can be compared to identify seasonal patterns.</li>

<li>High sales do not always mean high profit.</li>

<li>Some categories and sub-categories may require pricing or discount review.</li>

<li>Regional and state-level charts help identify stronger and weaker markets.</li>

<li>Customer segment analysis helps show which customer groups drive revenue.</li>

</ul>

::: warning-point
The most important interpretation is to compare **sales and profit together**. A business area with high sales but low profit may need closer review.
:::
::::

### 6. Recommendations

::: report-card
Based on the dashboard design and the types of insights it supports, Superstore should:

<ul class="report-list">

<li>Track monthly sales and profit together instead of focusing only on revenue.</li>

<li>Investigate high-sales areas with low or negative profit.</li>

<li>Review discounts in low-profit categories and sub-categories.</li>

<li>Focus marketing and inventory planning on profitable regions and segments.</li>

<li>Use filters to compare performance across time, products, customers, and geography.</li>

</ul>
:::

## Row {data-height="360"}

### 7. Limitations

::: report-card
The dashboard is useful for exploration, but it has limits.

<ul class="report-list">

<li>The data are historical and do not prove causation.</li>

<li>The dashboard does not include every possible business cost.</li>

<li>Profit results depend on the accuracy of the profit field in the dataset.</li>

<li>Some patterns may be affected by discounts, shipping, or business rules not fully shown in the data.</li>

<li>The dashboard does not forecast future sales or profit.</li>

</ul>
:::

### 8. Conclusion

::: report-card
The dashboard provides a clear and interactive view of the Superstore performance. It combines KPIs, monthly sales and profit trends, product analysis, geographic comparisons, customer insights, and a searchable data table to analyze performace in the store. 
:::
